---
title: "Exploration of Central Limit Theorem using Exponential Distribution"
author: "Bradley JK Smith"
date: "25 August 2016"
output: html_document
---
## Summary
This is an investigation of the Central Limit Theorem using repeated draws from exponential distribution.  

The CLT states that the distribution of averages of iid variables becomes that of a standard normal as the sample size increases. Alternatively, X_n ~ N(mu, sigma^2/n) (ref - B Caffo, Little Inference Book, p. 57)

## Set up
Load required libraries  
Set seed so code is reproducible  
Define lambda, number of draws and number of simulations  
```{r setup}
suppressPackageStartupMessages({
  require(ggplot2)
  })
set.seed(92361)
lambda <- 0.2
n_draws <- 40
n_simulations <- 1000
```
## Background

Exponential distribution:  

* PDF(x, lambda) = lambda * exp( -lambda * x) (where x >= 0)  
* Mean = 1/lambda  
* SD = 1/lambda  

For lambda = 0.2, the probability density for the Exponential distribution is shown below. In this case, the mean is 5 and the standard deviation is 5.

```{r}
x <- seq(0, 20, by=0.1)
y <- dexp(x, lambda)
qplot(x,
      y,
      main = 'Probability Density for Exponential Distribution (lambda = 0.2)',
      ylab = 'Probability',
      ylim = c(0,0.2),
      geom = c('line')
     )
```


## Simulation

Take 1000 averages of 40 exponentials.

```{r}
y <- NULL
for( i in 1 : n_simulations ) y <- c(y, mean(rexp(n_draws,lambda)))
dat <- as.data.frame(y)
colnames(dat) <- 'value'
```

## Comparison
Expect that the sampled distribution should be ~ N(mu, sigma^2/n)  
In this case:  

* mu = 1/lambda = 5
* sigma = 1/lambda = 5
* sigma^2/n = 25/40

Wald confidence intervals calculated under normal distribution

```{r}
mu_hat <- mean(dat$value)
sigma_hat <- sd(dat$value)
mu <- 1/lambda
sigma <- 1/lambda/sqrt(n_draws)
ll <- mu - qnorm(0.975) * sigma
ul <- mu + qnorm(0.975) * sigma

ggplot(data = dat, aes(x=value)) +
  geom_histogram(binwidth = 0.175, colour = 'lightblue', fill = 'lightblue', aes(y = ..density..)) + 
  geom_vline(xintercept = mu, size = 0.5, colour = 'blue', linetype = 'dashed') +
  geom_vline(xintercept = mu_hat, size = 1) +
  geom_vline(xintercept = ul, size = 0.5, colour = 'blue', linetype = 'dashed') +
  geom_vline(xintercept = ll, size = 0.5, colour = 'blue', linetype = 'dashed') +
  ggtitle('Comparison of Sample Distribution with theoretical values') +
  stat_function(fun = dnorm, args = list(mean = mu, sd = sigma), size = 0.5)
```


The sample distribution has the same shape as the theoretical normal distribution (shown with the overlaid curve) (albeit a bit narrower and skewed to the left). The mean of the distribution (shown with the thicker line) is very close to the the theoretical value and well within the Wald confidence intervals (shown with the dotted lines).

```{r, echo=FALSE}
cat('Mean of distribution: ', mu_hat, ' compared to confidence interval [', ll, ',', ul, ']\n')
cat('SD of distribution: ', sigma_hat, ' compared to theoretical value: ', sigma, '\n')
```
