---
title: "Exploration of Tooth Decay Data"
author: "Bradley JK Smith"
date: "25 August 2016"
output: html_document
---
## Summary
Basic exploratory data analysis of ToothGrowth dataset was performed.  
The null hypothesis that OJ supplement has no effect on tooth growth was rejected at the 95% confidence level.  
The null hypothesis that VC supplement has no effect on tooth growth was rejected at the 95% confidence level.

Under the assumption that supplement dosages are comparable, the null hypothesis that the OJ and VC supplements has similar effect on tooth growth was not rejected at the 95% confidence level.

## Set up
Load required libraries and data  
```{r setup}
suppressPackageStartupMessages({
  require(ggplot2)
  })
data('ToothGrowth')
```

## Background
The ToothGrowth data set has 3 levels of doses (0.5, 1.0 and 2.0) and 2 levels of supplements (OJ and VC) (I assume these are Orange Juice and Vitamin C, but can't label these on the graphs as I don't know).  

```{r}
table(ToothGrowth$dose, ToothGrowth$supp)
```

The graph of dosage versus length for each supplement is show below.

```{r}
qplot(x = dose, y = len, data = ToothGrowth, facets = . ~ supp)
```

## Hypothesis Testing

A general assumption is that the data is un-paired (which seems sensible as there is no 'patient' index in the data)

### Dosage

The null hypothesis is that the tooth growth (specified as 'len') is not affected by the dosage for each supplement. This is tested by taking pair-wise combinations of the dosage sets and performing an ungrouped two-sided t test. The p-value is adjusted by controlling the false discovery rate.

### Supplements

The null hypothesis is that the tooth growth is not affected by the supplement used when the dosage is controlled. This is tested by taking pair-wise combinations of the supplements at the same dosage and performing an ungrouped two-sided t test. In addition, an adjusted p-value is calculated by controlling the false discovery rate. 

The assumption here is that the dosage values are comparable. That is, a 0.5 dose of the OJ supplement can reasonably be compared to a 0.5 dose of the VC supplement.

```{r}
res <- as.data.frame(matrix(NA,ncol=6, nrow=9))
res[,1] <- c(rep('OJ',3), rep('VC',3), rep('OJ',3))
res[,2] <- c(rep(c(0.5, 0.5, 1.0),2), c(0.5, 1.0, 2.0))
res[,3] <- c(rep('OJ',3), rep('VC',3), rep('VC',3))
res[,4] <- c(rep(c(1.0, 2.0, 2.0),2), c(0.5, 1.0, 2.0))
colnames(res) <- c('supp_A', 'dose_A', 'supp_B', 'dose_B', 'p-value', 'p-adjust')

for( i in 1:nrow(res) ) {
  idx1 <- ToothGrowth[,'supp'] == res[i,1]
  idx2 <- ToothGrowth[,'dose'] == res[i,2]
  idx3 <- ToothGrowth[,'supp'] == res[i,3]
  idx4 <- ToothGrowth[,'dose'] == res[i,4]
  res[i,5] <- t.test(ToothGrowth[idx1&idx2,'len'], ToothGrowth[idx3&idx4,'len'], alternative = 'two.sided')$p.value
  if( i %% 3 == 0) {
    res[(i-2):i,'p-adjust'] <- p.adjust(res[(i-2):i,'p-value'], method = 'BH')
  }
}
res[,5:6] <- round(res[,5:6],5)

print(res, row.names=FALSE)
```


## So what does this mean?

1. Reject the null hypothesis that the OJ supplement has no impact on tooth growth at the 95% confidence level  
  + The adjusted p-values for rows 1 to 3 are all < 0.05. This is in line with the graph.

2. Reject the null hypothesis that the VC supplement has no impact on tooth growth at the 0.05 confidence level

  + The adjusted p-values for rows 4 to 6 are all < 0.05 (much less in this case). This is also in line with the graph where there is almost no overlap between the dose distributions.

3. Fail to reject the null hypothesis that one supplement is better than the other for all dosage levels.

  + The adjusted p-values for rows 7 to 9 are not all < 0.05. You can't say that OJ always beats VC no matter the dose.

  + Taking the dosage levels separately, you can make the case (at the 95% confidence level) that OJ is better for tooth growth at lower dosage levels - but this is very dependent on the assumption that dosage levels can be compared.